The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-07-22 paper.tex
Added intro to Chaha and some 2C data

>>>>

Chaha is a Southern Semitic language spoken in central Ethiopia by about 440,000 people.  
As is common with Semitic languages, it has many underlying consonant clusters, some of which are resolved by epenthesis of the high central vowel [\textipa{1}].  Vowel epenthesis appears to preferentially target rising and level sonority clusters over falling ones.  The data in this section mostly comes from Rose 2000, who also provides an analysis in terms of Syllable Contact.

We will use the existing sonority scale given in \ref{standardsonorityscale}, with a small modification. Chaha has a bilabial approximant represented by [\textipa{B}] (more precisely, \textipa{\|`B}) \citep[15]{banksira.2000} that is not as sonorous as /r/, whose realisation is tap-like \citep[181]{taranto.2001}. The justification for /\textipa{B}/ being less sonorous than /r/ is largely empirical, based on the epenthetic behaviour of clusters containing this sonorant \citep{405}{rose.2000}, data for which will be presented later in this paper. I will specify its sonority as an intermediate 3.5, under the sonority of 4 for liquids in general.

\ex. Sonority scale to be used in Chaha

\begin{center}
\begin{tabular}{ccccccccccccc}
 V & $>$ & w j & $>$ & r & $>$ & \textipa{B} & $>$ & m n & $>$ & f s z x & $>$ & t t' k k' d g \\
 vowels & & glides & & liquids & &   & & nasals & & fricatives & & stops \\
    V   & &    G   & &    R    & & \textipa{B} & & N & & F & & T \\
    6   & &    5   & &    4    & &    3.5      & & 3 & & 2 & & 1 \\
\end{tabular}
\end{center}
\subsection{CC clusters}

Underlying two-consonant clusters in Chaha are obligatorily epenthesised into if initial, never broken up by epenthesis in medial position, while final clusters are epenthesised into if they are of rising or level sonority, or the cluster consists of [r] followed by a nasal.

\ex. No onset clusters in Chaha
     \a. \textipa{k1r@tam} `he lifted up'

\ex. Tautomorphemic medial two-consonant clusters are tolerated no matter what their sonority profile
     \a. \textipa{s1rto} `cauterise ({\sc masc pl})!'
     \b. \textipa{n1k'mo} `pick ({\sc masc pl})!'
     \b. \textipa{at'met'} `solidified juice from [\textipa{@s@t}] plant'
     \z. \citep[(1a,c,d)]{rose.2000}

The presentation of the patterns of epenthesis in CC\# clusters is slightly complicated by variability between idiolects.  Epenthesis is possible in all level and rising sonority clusters.  
Among these clusters, the ones for which epenthesis is optionally unavailable are the obstruent-obstruent clusters.  Most falling sonority clusters do not undergo epenthesis,
except [r]-sonorant clusters, which can be broken up in certain idiolects. Voicing is irrelevant
to whether clusters undergo epenthesis.

The following table summarises the data.

\ex. Table showing epenthesis patterns for clusters of varying sonority patterns:

\begin{tabular}{c | c c c c c}
 \backslashbox{C1}{C2}      & Stop & Fricative & Nasal & Liquid \\ \hline
Stop   & \textbf{OPTIONAL}   & \textbf{OPTIONAL} & \textbf{\tickYes} & \textbf{\tickYes} \\
       & \textipa{n1g(1)d} & \textipa{d1g(1)s} & \textipa{n1k1m} & \textipa{g1d1r} \\
       & `touch!' & `give feast!' & `pick!' & `put to bed!' \\ \hline
Fric-  & \textbf{\tickNo} & \textbf{OPTIONAL} & \textbf{\tickYes} & \textbf{\tickYes} \\
ative  & \textipa{k1ft}   & \textipa{mes(1)x} &    & \textipa{s1f1r} \\
       & `open!'  & `chew!' & & `measure!' & \\ \hline
Nasal  & \textbf{\tickNo}     & \textbf{\tickNo}     & \textbf{\tickYes} & \textbf{\tickYes} \\
       & \textipa{g1nd} & \textipa{t1mx} & \textipa{g@n1m}   &  \\
       & `log (noun)'   & `scoop with finger!' & `take back loaned cow!' &  \\ \hline
Liquid & \textbf{\tickNo}     & \textbf{\tickNo} & \textbf{OPTIONAL}      & \textbf{\tickYes} \\
       & \textipa{f1rt} & \textipa{t1rx} & \textipa{k1r(1)m} & \textipa{s1B1r} \\
       & `divide in half!' & `make incision!' & `insult!'    & `break!' \\  \hline
\end{tabular}	
\noindent (Data assembled from \citet[404--7]{rose.2000})
